Practical consistency management for geographically distributed MMOG servers

### 2011-04-14 textos/optpaxos/draft.tex

The basic idea of the optimistic ordering is the following: assuming that the nodes have a synchronized clock\footnote{We don't require here perfectly synchronized clocks, as the optimistic protocol tolerates mistakes by its very definition. We only need clocks which are synchronized enought, so that our delivery order prediction succeeds and matches the conservative delivery order.}, whenever a server (either coordinator or replica) receives a command $C$ from a client, it immediately applies a provisory (optimistic) timestamp $pTS$ for it, which consists of the current time $now$ plus the estimated communication delay $dc(S)$ between $S$ and the coordinator. Therefore, $pTS~=~now~+~dc(S)$. A wait window of length $w(S)$ is considered, where $w(S)$ is defined as the highest estimated communication delay between the server $S$ and any of the other servers of its group. We will come back to $dc(S)$ later.

After applying the provisory timestamp to $C$, the server immediately forwards it to the other servers of that group, and puts $C$ on a list, where it stays until $now > C.pTS + w(S)$, which means that $C$ has been in that list for a time longer than the defined wait window. In the meantime, other commands, sent from other clients and forwarded by other servers, may have been received and also inserted in that list, always sorting by the provisory timestamp of these commands. If $w(S)$ has been correctly estimated as the highest communication delay between the server and each of the other servers, and no message was lost, than all the commands that were supposed to be delivered before $C$ have necessarily been received already. If the same occurred for all the servers, then all of them have received all the commands, and can deliver them in the same order of $pTS$.

However, even if the optimistic order is the same for all the servers, it won't be valid if the conservative order is different from it. That is why we include the value $dc(S)$ in the provisory timestamp of each command forwarded by $S$. Assuming that the clocks are sufficiently synchronized and that the communication delays have been correctly estimated, then if $C$ was sent by $S$ to the coordinator at the instant $t$, it will be received at the instant $t + dc(S)$. Thus, when two servers $S_1$ and $S_2$ forward the commands $C_1$ and $C_2$ to the coordinator and we set their timestamps as $pTS = now + dc(S)$, then the coordinator will receive\footnote{It's very unlikely that provisory timestamps won't clash, but it is reasonable to assume that, for commands affecting the same objects, and given a timestamp resolution of a few miliseconds, they will probably have sufficiently different timestamps, so that the receival order at the coordinator matches the one predicted by its replicas.} $C_1$ before $C_2$ iff $C_1.pTS~<~C_2.pTS$. Therefore, given the optimistic assumptions we have made, the conservative delivery order (based on the order in which the coordinator received the commands) will be the same as the optimistic delivery order (based on the provisory timestamp assigned by the replicas).

@@

Unfortunately, even with a very good delay estimation (e.g. on an environment with a low jitter), there is absolutely no guarantee that the optimistic delivery order will match the conservative one. When it doesn't, it is considered a mistake. Every mistake of the optimistic algorithm -- either a lost command message, or an out-of-order delivery -- will cause a rollback of the optimistic state of the objects and re-execution of some of the optimistically delivered commands.

To perform that, we consider that each object has an optimistic delivery queue, $Q_{opt}$. Whenever a command is optimistically delivered, the optimistic state is updated and the command is pushed in the back of $Q_{opt}$. Whenever a command $C_c$ is conservatively delivered, it updates the conservative state of the object and the algorithm checks whether it is the first command in $Q_{opt}$. If it is, $C_c$ is simply removed from $Q_{opt}$ and the execution continues. If it isn't, it means that a command was either lost or was optimistically delivered out of order. We then check whether $Q_{opt}$ contains $C_c$. If it does, it is removed from that list -- if $Q_{opt}$ doesn't contain $C_c$, it was probably lost, or might be the very unlikely case where the conservative delivery happened before the optimistic one, so $C_c$ is stored in a list of possibly delayed optimistic delivery and, if it is ever optimistically delivered, the algorithm will know that it should only discard that command. Then, the optimistic state is overwritten with the conservative one and, from that state, all the remaining comands in $Q_{opt}$ are re-executed, leading to a new optimistic state for that object.

>>>>

The basic idea of the optimistic ordering is the following: assuming that the nodes have a synchronized clock\footnote{We don't require here perfectly synchronized clocks, as the optimistic protocol tolerates mistakes by its very definition. We only need clocks which are synchronized enought, so that our delivery order prediction succeeds and matches the conservative delivery order.}, whenever a server (either coordinator or replica) receives a command $C$ from a client, it immediately applies a provisory (optimistic) timestamp $pTS$ for it, which consists simply of the current value of the server's wallclock, $now$. Therefore, $pTS=now$. A wait window of length $w(S)$ is considered, where $w(S)$ is defined as the highest estimated communication delay between the server $S$ and any of the other servers of its group.

After applying the provisory timestamp to $C$, the server immediately forwards it to the other servers of that group, and puts $C$ on a list, where it stays until $now>C.pTS+w(S)$, which means that $C$ has been in that list for a time longer than the defined wait window. In the meantime, other commands, sent from other clients and forwarded by other servers, may have been received and also inserted in that list, always sorting by the provisory timestamp of these commands. If $w(S)$ has been correctly estimated as the highest communication delay between the server and each of the other servers, and no message was lost, than all the commands that were supposed to be delivered before $C$ have necessarily been received already. If the same occurred for all the servers, then all of them have received all the commands, and can deliver them in the same order of $pTS$.

However, even if the optimistic order is the same for all the servers, it won't be valid if the conservative order is different from it. For that reason, before submitting a command to be conservatively ordered, the coordinator $S_c$ also waits for $w(S_c)$, which is the maximum delay between $S_c$ and any of its replicas. Assuming that the clocks are sufficiently synchronized and that the communication delays have been correctly estimated, then after the time $w(S_c)$ has elapsed, the coordinator will have received all the commands forwarded by any of its replicas (until the instant $t:t=now-w(S_c)$), and it will be able to conservatively order them in the same order of the optimistic delivery, assuming that the final order is the same as the order in which the messages were submitted by the coordinator (FIFO). Therefore, given the optimistic assumptions we have made, the conservative delivery order (based on the order in which the coordinator FIFO-broadcast the commands to the group) will be the same as the optimistic delivery order (based on the provisory timestamp assigned by the replicas).

@@

Unfortunately, even with a very good delay estimation (e.g. on an environment with a low jitter), there is absolutely no guarantee that the optimistic delivery order will match the conservative one. When it doesn't, it is considered a \emph{mistake}. Every mistake of the optimistic algorithm -- either a lost command message, or an out-of-order delivery -- will cause a rollback of the optimistic state of the objects and re-execution of some of the optimistically delivered commands.

To perform that, we consider that each object has an optimistic delivery queue, $Q_{opt}$. Whenever a command is optimistically delivered, the optimistic state is updated and the command is pushed in the back of $Q_{opt}$. Whenever a command $C_c$ is conservatively delivered, it updates the conservative state of each object in $obj(C_c)$ and, for each one of them, the algorithm checks whether it is the first command in $Q_{opt}$. If it is, $C_c$ is simply removed from $Q_{opt}$ and the execution continues. If it isn't, it means that $C_c$ was either optimistically delivered out of order, or it was simply never optmistically delivered. It then checks whether $Q_{opt}$ contains $C_c$. If it does, it means the command was optimistically deliverd out of order, and it is removed from the list -- if $Q_{opt}$ doesn't contain $C_c$, it was probably lost\footnote{Also, when $C_c$ is delivered, but it is not in $Q_{opt}$, the remaining possibility is the very unlikely case where the conservative delivery happened before the optimistic one. To handle this case, $C_c$ is stored in a list of possibly delayed optimistic delivery and, if it is ever optimistically delivered, the algorithm will know that it should only discard that command, instead of updating the optimistic state.}. Then, the optimistic state is overwritten with the conservative one and, from that state, all the remaining comands in $Q_{opt}$ are re-executed, leading to a new optimistic state for that object.
